---
title: "mydata dataset - Data Exploration and Cleanup"
format: 
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error= TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Load Data

Read in the CSV file using the `readr` package

```{r}
library(readr)
mydata <- read_csv("mydata.csv")
```

## Take a look at the dataset

Type `View(mydata)` at the prompt or run interactively from an R script or Rmarkdown document.

I will also be handing out a hard copy of these data so you can look at the data, the values and any patterns that emerge.

```{r eval=FALSE}
View(mydata)
```

```{r echo=FALSE}
library(DT)
datatable(
  mydata,
  options = list(
    pagelength = 21,
    autowidth = TRUE
  )
)
```


## Also take a look at the "codebook"

Feel free to click on links below or right click and "SaveAs" to save the file on your computer.

* [Codebook in DOC format](Mydata_Codebook.docx)
* [Codebook in PDF format](Mydata_Codebook.pdf)


## Open discussion on what you "see"

There are 11-12 problems with this little dataset. Which ones do you see?

1. How might we use code to FIND (discover) these problems?

2. How might we use code to FIX (correct) these problems?

## Code to find and fix these issues

[link to be added later](https://melindahiggins2000.github.io/mydata/)